---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

### Author: Wellintton Perez 2018

Loading some packages we will need later
```{r results="hide"}
warn.conflicts = FALSE
library(dplyr,quietly=TRUE)
library(ggplot2,quietly=TRUE)
```

## 1. Loading and preprocessing the data
```{r echo=TRUE}
location=getwd() #use the current working directory

url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
filename<-"activity.csv"

data_location<-file.path(location,"data") # creates a repository to keep the raw data

data_file<-file.path(data_location,filename) #concat path and file name

#if the directory data does not existing download fresh data
if(!file.exists(data_file)){
  message("Download the activity data set.")
  rawdata<-tempfile()
  rawdata<-paste(rawdata,".zip")
  download.file(url,rawdata) #download data from the URL
  unzip(rawdata,exdir=data_location)
}

data=read.csv2(data_file,sep=",") #Reading the file

data$date=as.Date(data$date)  #need to convert to date

```
## What is mean total number of steps taken per day?

calculating the number of steps taken per day
```{r}

data_with_no_na<-na.omit(data) #removing missing values 

#group by date to summarize the data
by_date<-group_by(data_with_no_na,date) 

steps_per_day<-summarize(by_date,
                         obs=n(),
                         total_steps=sum(steps),
                         average_steps=mean(steps))
```
### Histogram of the total number of steps taken each day
```{r}
#histogram
ggplot(steps_per_day,aes(color=date))+geom_histogram(aes(total_steps))+
  ggtitle("Histogram of total steps per day")+
  xlab("Total steps")+
  ylab("Frequency")
```

### This is a table containing the calculated mean and median steps per day
```{r xtable,results="asis"}
library(xtable,quietly=TRUE)
table_avg_steps<-xtable(steps_per_day)
print(table_avg_steps,type="html")
```

## What is the average daily activity pattern?
### 1. Time series plot showing the number of steps taken on average across all days
```{r}
average_steps_ts<-ts(steps_per_day$average_steps,frequency = 1)
plot.ts(average_steps_ts,xy.labels=c("Time series","Average steps per day"))
title("Average number of steps taken per day")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
highest<-data[order(-data$steps),]
head(highest) #top 6 steps per 5 minute intervals
```
The 5-minute interval with the highest number of steps if interval `r highest[[3]][1]` with `r highest[[1]][1]` steps. 

## Imputing missing values
### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs) 
```{r}
temp_NA<-is.na(data)
missing_values<-length(data[temp_NA])
```
####The total number of missing values in the dataset is `r missing_values`.

### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

After careful analysis of the dataset I decided to set all missing values to zero.  The reason is that the complete day was missing values so I could not find a good method to impute the missing values.

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
warn.conflicts = FALSE
data[is.na(data$steps),][[1]]<-0.001 #fill in missing values with zero

#group by date to summarize the data
by_date<-group_by(data,date) 

steps_per_day<-summarize(by_date,
                         obs=n(),
                         total_steps=sum(steps),
                         average_steps=mean(steps),
                         median_steps=median(steps))
```
### 4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
#histogram
ggplot(steps_per_day,aes(color=date))+geom_histogram(aes(total_steps))+
  ggtitle("Histogram of total steps per day")+
  xlab("Total steps")+
  ylab("Frequency")
```

### This is a table containing the calculated mean and median steps per day after imputing the missing values
```{r xtable2,results="asis"}
library(xtable,quietly=TRUE)
table_avg_steps<-xtable(steps_per_day)
print(table_avg_steps,type="html")
```

#### Do these values differ from the estimates from the first part of the assignment? 
#### What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
#simple function to determine the weekend from weekday
isWeekDay<-function(days){
  temp<-c()
  for(day in days){
    if(day=='Saturday'|day=='Sunday')
      temp<-c(temp,"weekend")
    else
      temp<-c(temp,"weekday")
  }
  return(temp)
}

data_day<-mutate(steps_per_day,weekday=factor(isWeekDay(weekdays(steps_per_day$date)))) #mutate to add a new column with the week day

```
### Panel plot containing the time series 
```{r}
xyplot(total_steps ~ weekday|weekday, 
           data = data_day,
           type = "l",
           xlab = "Interval",
           ylab = "Number of steps",
           layout=c(1,2))
```